### 2012-09-03 paper_sean/conclusion.tex
JMC edits
editting, cutting paper

HyperNEAT has proven to be an effective encoding in gait evolution and is able to find gait regularities without human intervention \cite{clune1}, \cite{clune2}, \cite{yos:clune}. 
Glette et al. \cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware.  
This was possible because the simulator afford a much larger number of evaluations to be performed during the evolution process. 
In this study, we took the two -- a sophisticated algorithm (HyperNEAT) and a physical simulator -- to see how HyperNEAT would perform with a many evaluations in the evolution process. 
We showed that HyperNEAT, when used with a simulator, could evolve very impressive gaits. 
In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\%. 
In the real life trials, HyperNEAT's best gait performed 5.4\% better than the GA gait. 
This study also showed that HyperNEAT will exploit the characteristics of the problem in order to maximize its fitness.
Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
However, because of the physical limitations of the robot, these exploitations could not be transferred well to reality. 
Moreover, the inevitable faults with the simulator (it cannot model the robot or the real world perfectly) added to the difficulty in the transfer from simulation to reality. 
Thus, many of the gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.

Next steps for this research heads towards improving the transferabilities of the gaits from simulation to reality. 
The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
Many different approaches to reduce the reality gap has been proposed \cite{koos1}, \cite{bongard}, \cite{zagal}.
To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 
The one easiest and most promising to implement is the Transferability Approach \cite{koos2} which has shown to be successful in quadrapedal gait evolution. 
HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait on the QuadraTot platform.

>>>>

% !TEX root =main.tex
HyperNEAT has proven to be an effective encoding in gait evolution which is able to find gait regularities without human intervention \cite{clune1,clune2,yos:clune}. 
Glette et al. \cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware.  
This was possible because the simulator afford a much larger number of evaluations to be performed during the evolution process. 
In this study, we took the two -- a sophisticated algorithm (HyperNEAT) and a physical simulator -- to see how HyperNEAT would perform with a many evaluations in the evolution process. 
We showed that HyperNEAT, when used with a simulator, could evolve very impressive gaits. 
In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\%. 
In the real life trials, HyperNEAT's best gait performed 5.4\% better than the GA gait. 
This study also showed that HyperNEAT will exploit the characteristics of the problem, such as ferquency and regularity, in order to maximize its fitness.
%Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
%When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
However, because of the physical limitations of the robot, these exploitations did not transfer well to reality. %could not be transferred well to reality.
Moreover, the inevitable faults with the simulator (it cannot model the robot or the real world perfectly) added to the difficulty in the transfer from simulation to reality. 
Thus, many of the gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.

Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality. 
The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
Many different approaches to reduce the reality gap has been proposed \cite{koos1,bongard,zagal}. 
To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 
The one easiest and most promising to implement is the Transferability Approach \cite{koos2} which has shown to be successful in quadrapedal gait evolution. 
HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait on the QuadraTot platform.

### 2012-09-03 paper_sean/hyperneat.tex
Updated figure commands, added as many figures as possible.
Added hyperneat_bug_example figure to Sean's paper
JMC edits
editting, cutting paper

HyperNEAT is a neuroevolution method which evolves artificial neuron networks (ANNs) using evolutionary algorithms.
It indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
HyperNEAT has shown that it could evolve large neural networks that represent the networks and features present in brains such as repetitions and regularities \cite{stanley3}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson}. 
It has been shown that objects evolved using HyperNEAT look complex, natural and feature characteristics such as symmetries and repetition of themes which are often found in nature. % figure from jeff paper

The motivation for CPPNs is that complex patterns can be produced by determining attributes of their phenotypic components as a function of their geometric location. 
This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.

Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. % figure!
Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop.
For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
These functions allow evolution to exploit various properties such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
Because of evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.

@@

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=CPPN.eps, width=8cm}
\caption [ ]{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2}.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=hyperneat.eps, width=8cm}
\caption [ ]{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2}}.
\end{center}
\end{figure}

>>>>

% !TEX root =main.tex
HyperNEAT is a neuroevolution method which evolves artificial neuron networks (ANNs) using evolutionary algorithms.
HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.

HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
The motivation for CPPNs is that complex patterns can be produced by determining attributes of their phenotypic components as a function of their geometric location. 
This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.

Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop.
For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
These functions allow evolution to exploit various properties such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
Because of evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.

@@

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=CPPN.eps, width=8cm}
% \caption [ ]{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2}.}
% \end{center}
% \end{figure}
\figp{hyperneat_bug_example}{.6}{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2}.}

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=hyperneat.eps, width=8cm}
% \caption [ ]{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2}}.
% \end{center}
% \end{figure}
\figp{hyperneatExplanation}{.6}{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2}}

### 2012-09-03 paper_sean/methods.tex
JMC edits
editting, cutting paper

Gaits were first evolved in simulation, as previous done in Glette et al. \cite{glette}. 
During the evolution process, we evolved 200 organisms for 200 generations in 20 runs with different initial conditions specified by randomly seeding the run. 
To keep in line with previous studies that have been performed on the QuadraTot platform, each gait ran for 12 seconds. 
All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al. \cite{yos:clune} except for the frequency of the sine wave function used. 
Through experimentation, we found that high frequency gaits, gaits that require servos to change their directions often, were much more prone to servo shutdowns than low frequency gaits (at around 1.5Hz per servo). 
HyperNEAT often evolved high frequency gaits (over 8Hz per servo) which performed poorly in reality due to servo constraints. 
This led to gaits performing dramatically different in real-life compared to simulations. 
To address this issue, two changes were made:
First, the frequency of the sine wave function used in HyperNEAT was reduced by a factor of 6.63 by changing the sine function to

@@

from

\begin{equation}
sin(t/6)\pi
\end{equation}

so that HyperNEAT would lean towards evolving low frequency gaits. Sine frequency of 0.004  was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
Second, a policy that punished high frequency gaits was implemented during the evolution process. 
If the average frequency of each for the gait was over the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:

@@

Through experimentation, we chose the threshold frequency to be 1.67 Hz/servo. After implementing the frequency punishment policy, the servos shutdown much less frequently.

@@

where \emph{d} is the distance the center of mass of the robot travelled during the given time as was done in Clune et al. \cite{clune1}.
The distance travelled was calculated as the Euclidean distance travelled by the robot in the XY plane during the alotted time, in centimeters.

In simulation, the distance travelled was calculated using the output file generated by the simulator.
In real-life, the distance travelled was calculated using the log file recorded using the Wiimote system described earlier in the paper. For real-life trials, the actual fitnesses were not calculated and the gaits were evaluated using only the velocity of the robot.

To evolve gaits using the physical simulator, we wrote a Python script\footnote{All scripts and drivers available at http://quadratot.yosinski.com} that interfaced between the simulator and HyperNEAT which allowed us to evolve gaits fast and efficiently with no human intervention.
The runs were performed on three separate machines all running the same version of Linux (Ubuntu 11.04) using the same HyperNEAT settings and files to speed up the simulation process.

>>>>

% !TEX root =main.tex
Gaits were first evolved in simulation and then transferred onto the real robot, as previous done in Glette et al. \cite{glette}. 
During the evolution process, we evolved 200 organisms for 200 generations in 20 runs with different initial conditions specified by randomly seeding the run. 
Each gait was 12 seconds long, both in simulation and in reality. 
All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al. \cite{yos:clune} except for the frequency of the sine wave function used, which was changed in order to address the servo shutdown problem mentioned previously. 
We found that gaits that require servos to change their directions often were more prone to servo shutdowns than low frequency gaits. %(at around 1.5Hz per servo). 
But with the sine wave frequency used for Yosinski et al. \cite{yos:clune}, HyperNEAT often evolved high frequency gaits (over 8Hz per servo) which performed poorly in reality due to servo constraints. 
This led to gaits performing dramatically different in real-life compared to simulations. 
%To address this issue, two changes were made:
To address thsi, we reduced by a factor of 6.63 by changing the sine function to

@@

%from
%
%\begin{equation}
%sin(t/6)\pi
%\end{equation}
%

so that HyperNEAT would lean towards evolving low frequency gaits. %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
We further added a policy that punished high frequency gaits while evolving in simulation. 
If the average frequency of each leg for the gait was higher than the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:

@@

We chose the threshold frequency for punishment to be 1.67 Hz/servo. 
After implementing the frequency punishment policy, the servos shutdown much less frequently.

@@

where \emph{d} is the Euclidean distance the center of mass of the robot travelled during the gait duration. %as was done in Clune et al. \cite{clune1}.

%In simulation, the distance travelled was calculated using the output file generated by the simulator.
%In real-life, the distance travelled was calculated using the log file recorded using the Wiimote system described earlier in the paper. For real-life trials, the actual fitnesses were not calculated and the gaits were evaluated using only the velocity of the robot.

To evolve gaits using the physical simulator, we wrote a Python script\footnote{All scripts and drivers available at http://quadratot.yosinski.com} that interfaced between the simulator and HyperNEAT. % which allowed us to evolve gaits fast and efficiently with no human intervention.
The runs were performed on three separate machines all running Ubuntu 11.04 using the same HyperNEAT settings and files to speed up the simulation process.

### 2012-09-03 paper_sean/results.tex
JMC edits
editting, cutting paper

HyperNEAT, used with a physical simulator, produced gaits that outperformed all previous gaits tried and recorded on the QuadraTot. 
The fastest gait evolved by HyperNEAT in simulation and transfered onto the real robot recorded a speed of 14.5 cm/s, faster than the previously fastest gait published in Glette et al. \cite{glette}. 
In their paper, Kyrre et al. \cite{glette} recorded 17.8 cm/s in their robot in Norway \cite{glette}, but to make sure that there were no differences in robots, the same gait was also tested on the robot used for our experimental testing.
The same gait, ran by using the gait file sent to us by Kyrre Glette, recorded an average speed of 12.951 cm/s and a maximum of 13.76 cm/s on the same QuadraTot platform in the Cornell Creative Machines Lab where Yosinski et al. \cite{yos:clune} and this study was done.
HyperNEAT, used with a simulator, outperformed Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run with 200 organisms and 200 generations as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run with 200 organisms and 300 generations.
The best gait evolved using HyperNEAT with simulator performed 49.5\% better than the best gait evolved using HyperNEAT in hardware \cite{yos:clune}. 
This is most likely due to the difference in the number of evaluations between the two studies. 
The previous study, HyperNEAT in hardware, only had 180 evaluations while in this study, a simulator allowed a much higher number -- 20000 -- of evaluations. 
Another minor reason for this may be that the implementation placed to reduce gait frequency may have helped in producing gaits that were less susceptible to shutdowns, as Yosinski et al. \cite{yos:clune} did not have such things.
Although most gaits started off its evolution process by producing fast and high frequency gaits, with more generations, the gaits dramatically reduced its frequencies while continually increasing their fitness. The average frequency of the final gaits was 1.29 Hz per servo.

Gaits produced in this study confirmed that with the use of a physical simulator, HyperNEAT is able to fully exploit the geometry of the problem without human intervention, which was not evident in gait evolution using HyperNEAT in hardware \cite{yos:clune}. 
HyperNEAT took full advantage of the regularity of the problem and produced gaits more regular, coordinated, and effective than all other gaits evolved for the QuadraTot platform in the past.
This is evident in the servo position plots where the gaits produced in this study have smooth and repeating curves while gaits produced by HyperNEAT in hardware produced much more ragged and random curves.
The difference in regularity between the two studies is most likely due to the number of evaluations that were done in gait evolution. Additionally, the noise in the real world was extremely high, making learning in hardware difficult, which may have further impeded the search for higher-performing, regular gaits.

@@

% table!
\begin{table}
\caption{The number of evaluations per run (gen x orgs), simulated and real trial velocities in [cm/s]. Data used from Yosinksi et al. \cite{yos:clune}, Glette et al. \cite{glette}, and Shen et al. \cite{haocheng}.
*17.8 cm/s was recorded on Glette et al. \cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
\hline
                                         & Num. Evaluations  & Sim. Velocity  & Real. Velocity \\
\hline
Best parameter optimization learning    &153    & --    & 4.64 \\
\hline
Hardware HyperNEAT                  & 180         & --         &   9.7     \\
\hline
Simulated GA              & 60000       & 16.4       &   12.951*     \\
\hline
RL PoWER Spline                          & 300         & --         &   11.05 \\
\hline
Simulated HyperNEAT                      & 40000       & 25.4       &   14.5 \\
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

% !TEX root =main.tex
HyperNEAT, used with a physical simulator, produced gaits that outperformed all previous gaits tried and recorded on the QuadraTot.\footnote{Videos of the gaits in simulation and on robot available at http://quadratot.yosinski.com}
The speed of the fastest gait recorded in this study was 14.5 cm/s, faster than the previously fastest gait published in Glette et al. \cite{glette}. 
Kyrre et al. \cite{glette} recorded 17.8 cm/s on their robot, but to make sure that there were no differences in robots, the same gait was also tested on the robot used for this study (which was the one used by Yosinski et al. \cite{yos:clune} and Shen et al. \cite{haocheng}. 
The same gait recorded an average speed of 12.951 cm/s and a maximum of 13.76 cm/s on our robot. We used these numbers in comparing results. 
The gait from this study outperformed the previous Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run.
It also outperformed the best gait evolved using HyperNEAT in hardware \cite{yos:clune} by 49.5\%, most likely due to the difference in the number of evaluations between the two studies -- 40000 vs 180. 
Another minor reason for superior performance may be that frequency limiter policy that we placed may have helped in producing gaits that were less susceptible to shutdown while Yosinski et al. \cite{yos:clune} did not have such things.
Although most gaits started off its evolution process by producing fast and high frequency gaits, with more generations, the gaits dramatically reduced its frequencies while continually increasing their fitness. The average frequency of the final gaits was 1.29 Hz per servo.

%Gaits produced in this study confirmed that with the use of a physical simulator, HyperNEAT is able to fully exploit the geometry of the problem without human intervention, which was not evident in gait evolution using HyperNEAT in hardware \cite{yos:clune}. 
HyperNEAT took advantage of the regularity of the problem and produced gaits more regular, coordinated, and effective than all other gaits evolved for the QuadraTot platform in the past.
This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos. 
%The difference in regularity between the two studies is most likely due to the number of evaluations that were done in gait evolution. Additionally, the noise in the real world was extremely high, making learning in hardware difficult, which may have further impeded the search for higher-performing, regular gaits.

@@

% table!
\begin{table}
\caption{The number of evaluations per run (gen x orgs), simulated and real trial velocities in [cm/s]. Data used from Yosinksi et al. \cite{yos:clune}, Glette et al. \cite{glette}, and Shen et al. \cite{haocheng}.
*17.8 cm/s was recorded on Glette et al. \cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
\hline
                                         & Num. Evaluations  & Sim. Velocity  & Real. Velocity \\
\hline
Best parameter optimization learning    &153    & --    & 4.64 \\
\hline
Hardware HyperNEAT                  & 180         & --         &   9.7     \\
\hline
Simulated GA              & 60000       & 16.4       &   12.951*     \\
\hline
RL PoWER Spline                          & 300         & --         &   11.05 \\
\hline
Simulated HyperNEAT                      & 40000       & 25.4       &   14.5 \\
\hline
\end{tabular}
\end{center}
\end{table}
%JMC All numbers in the same column should be to the same precision (e.g. for the real velocity column, one decimal place). Also, can you right-justify the numbers (while still keeping them roughly centered)? Comparing numbers that are not right-justified is hard...so if you can't center and right-justify, then I let's see if just having it be right justified is better.

### 2012-09-03 paper_sean/simulator.tex
editting, cutting paper

% !TEX root =main.tex
Simulations occurred in C++ software customized for the QuadraTot \footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously~\cite{glette}. 
Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
%The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
%The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.%JMC: Assumed
%The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. %JMC: Assumed
%The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. %JMC: Assumed
%In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. %Do all results in this paper reflect this?
Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%

>>>>

% !TEX root =main.tex
Simulations occurred in C++ software customized for the QuadraTot\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously \cite{glette}. 
Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
%The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
%The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.%JMC: Assumed
%The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. %JMC: Assumed
%The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. %JMC: Assumed
%In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. %JMC/SL - Do all results in this paper reflect this? / They will soon.
Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%
